---
title: "Create ProjectDB SQLite Database"
author: "Larkin, Yvonne"
subtitle: 'Assignment 08.1: SQL Querires'
date: "Sp 2025, March 11"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r}
library(RSQLite)

fpath = ""
dbfile = "projectdb3.db"

# if database file already exists, we connect to it, otherwise
# we create a new database
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath,dbfile))
```

```{sql connection=dbcon}
drop table if exists projectmgr
```

```{sql connection=dbcon}
drop table if exists projects
```

```{sql connection=dbcon}
drop table if exists certifications
```

```{sql connection=dbcon}
drop table if exists earnedcerts
```

```{sql connection=dbcon}
create table projectmgr (
  pmid integer,
  name text not null,
  primary key (pmid)
)
```

```{sql connection=dbcon}
create table certifications (
  cid integer not null,
  cert text not null,
  body text not null,
  primary key (cid)
)
```

```{sql connection=dbcon}
create table earnedcerts (
  pmid integer,
  cert integer not null,
  primary key (pmid,cert)
  foreign key (pmid) references projectmgr(pmid)
  foreign key (cert) references certifications(cid)
)
```

```{sql connection=dbcon}
create table projects (
  pid integer,
  pname text default "No Name",
  budget double default 0,
  pmgr integer  not null,
  primary key (pid),
  foreign key (pmgr) references projectmgr(pmid)
)
```

```{sql connection=dbcon}
insert into certifications values
  (1,"PMP","Project Management Institute"),
  (2,"APM","Project Management Institute"),
  (4,"CBAP","International Institute of Business Analysis"),
  (6,"PSM","Scrum Organization"),
  (5,"CIPM","American Academy of Project Management")
```

```{sql connection=dbcon}
insert into projectmgr values
  (11,"Xi Xufen"),
  (22,"Ann Kelsey"),
  (18,"Sandeep Ranghani"),
  (19,"Alicia Knauss"),
  (38,"Roxanna da Silva"),
  (23,"Wagner Ormond"),
  (47,"Margaret O'Shea"),
  (43,"Kamala Patel"),
  (20,"Alden White")
```

```{sql connection=dbcon}
insert into earnedcerts values
  (11,1),
  (11,4),
  (18,1),
  (22,1),
  (19,1),
  (47,1),
  (23,1),
  (20,1),
  (20,4),
  (38,5)
```

```{sql connection=dbcon}
insert into projects values
  (100,"GOTAM",350000,11),
  (200,"Data Twist",875000,20),
  (300,"Alpha Launch",12500,18),
  (302,"Alpha Arch Design",38000,18),
  (303,"Alpha Dev",125000,18),
  (305,"Alpha Analytics Dashboard",22900,38),
  (505,"Training Portal Design",49900,19),
  (811,"Bootstrap Demo",0,47)
```

```{sql connection=dbcon}
select * from projectmgr
```

```{sql connection=dbcon}
select * from projects
```

```{sql connection=dbcon}
select * from certifications
```

```{sql connection=dbcon}
select * from earnedcerts
```
## ERD of Project Database
The ERD for the project database is shown below:

My [Lucid Chart]("https://lucid.app/documents/embedded/bd3e86ec-fb83-4592-8bbd-3b9d90adb7d5?invitationId=inv_a3e6a677-196a-4f84-8dac-d3e797a0a77b#") is giving me difficulty to embed in a pdf with the provided HTML link from Lucid. Therefore I am embedding the image for the pdf as well as providing the link for HTML.

![]("C:\Users\yvonn\Downloads\lucidchart"){width="70%"}

<div style="width: 640px; height: 480px; margin: 10px; position: relative;"><iframe allowfullscreen frameborder="0" style="width:640px; height:480px" src="https://lucid.app/documents/embedded/bd3e86ec-fb83-4592-8bbd-3b9d90adb7d5" id="zOt0-mHFiIS."></iframe></div>

### Q2.1 List the names of all project managers
```{sql connection= dbcon}
SELECT DISTINCT name
FROM projectmgr
ORDER BY name;
```
### Q2.2 How many projects mangers are there?
```{sql connection=dbcon}
SELECT DISTINCT count(name) AS 'Number of Managers'
FROM projectmgr;

```
### Q2.3 List the names of all project managers who manage at least one project, ordered alphabetically starting with 'A'
```{sql connection=dbcon}
SELECT M.name, COUNT(P.pid) AS 'Project Count'
FROM projectmgr M JoIN projects P ON (M.pmid = P.pmgr)
GROUP BY M.name
HAVING COUNT(M.name) > 0
ORDER BY M.name;
```
### Q2.4 List all project names, the name of the project manager, and the budget
```{sql connection=dbcon}
SELECT P.pname AS 'Project Name', M.name AS 'Project Manager', P.budget AS 'Project Budget'
FROM projectmgr M JoIN projects P ON (M.pmid = P.pmgr)
ORDER BY P.pname;
```
### Q2.5 What is the average budget for all projects?
```{sql connection=dbcon}
SELECT avg(budget) AS 'Average Budget Across All Projects'
FROM projects;
```
### Q2.6 Which project managers have more than two certificationa? List their names. This query requires a GROUP BY
```{sql connection=dbcon}
SELECT N.name, COUNT(C.cert) AS 'Certification Count'
FROM projectmgr N JoIN earnedcerts C ON (N.pmid = C.pmid)
GROUP BY N.name
HAVING COUNT(C.cert) > 2
ORDER BY N.name;
```
There are no managers with more than 2 certifications, if you change line 179 to 'HAVING COUNT(C.cert) >= 2' You will see that two managers (Alden White and Xi Xufen) have two certificates but no manager has above two.


### Q2.7 How many projects have a budget that is below the average budget?
```{sql connection=dbcon}
SELECT COUNT(pid) AS "Number of Projects below total average budget"
FROM projects
WHERE budget < (SELECT AVG(budget) FROM projects);
```
### Q2. Which Projects have a project name that contains the text "Design" anywhere in the project name. List names of the projects
```{sql connection=dbcon}
SELECT pname AS 'Project Name'
FROM projects
WHERE pname LIKE '%Design%'
```
### Q2.9 How many project managers manage projects that in total budget (i.e the sum of the budgets of their projects) are no more than $50,000?
```{sql connection=dbcon}
SELECT COUNT(*) AS 'Number of Project Managers'
FROM(
  SELECT M.name, SUM(P.budget) AS 'Total Budget'
  FROM projectmgr M JOIN projects P ON (M.pmid = P.pmgr)
  GROUP BY M.name
  HAVING SUM(P.budget) <= 50000
) AS subquery;
```

### Q2.10 Which project managers do not manage any projects? List their names
```{sql connection=dbcon}
SELECT name
FROM projectmgr
WHERE pmid NOT IN (SELECT pmgr FROM projects)
ORDER BY name;
```


```{r}
dbDisconnect(dbcon)
```

