---
title: "Analyze Data in CSV Using SQL"
subtitle: "Assignment 08.2"
author: "Larkin, Yvonne"
date: "Sp 2025, March 11"
output:
  html_document:
    df_print: paged
---


```{r setup, warning=FALSE, message=FALSE}
library(sqldf)
dbConcrete <- read.csv("http://s3.us-east-2.amazonaws.com/artificium.us/datasets/ConcreteStrengthData.csv")
```

### Q1 What is the average concrete strength?
```{r Q1}
sqlStmt <- paste0(
  "SELECT round(avg(strength), 2) AS 'Mean Strength Value for Concrete' ",
  "FROM dbConcrete ")
avg.str <- sqldf::sqldf(sqlStmt)
avg.str
```

### Q2 How many concrete samples were cured for more than 30 days?
```{r Q2}
sqlStmt <- paste0(
  "SELECT COUNT(age) AS 'Number of Samples Aged more than 30 days' ",
  "FROM dbConcrete ",
  "WHERE age > 30 ")
samples <- sqldf::sqldf(sqlStmt)
samples
```

### Q3 What was the average amount of aggregate (coarse and fine) used for concrete with a strength above 50?
```{r Q3}
sqlStmt <- paste0(
  "SELECT round(avg(coarse_agg), 2) AS 'Average Coarse Aggregate', round(avg(fine_agg), 2) AS 'Average Fine Aggregate' ",
  "FROM dbConcrete ",
  "WHERE strength > 50 "
)
avg_agg <- sqldf::sqldf(sqlStmt)
avg_agg
```

### Q4 How many concrete samples did not use any ash (i.e. had an ash content of 0)?
```{r Q4}
sqlStmt <- paste0(
  "SELECT COUNT(cement) AS 'Number of Concrete Samples with no ash' ",
  "FROM dbConcrete ",
  "WHERE ash = 0 ")
noAsh <- sqldf::sqldf(sqlStmt)
noAsh
```

### Q5 Define an outlier as any value that is more than 3 standard deviations from the mean. List the concrete samples (with all columns) that had a concrete strength that can be considered an outlier
```{r Q5}
#no standard deviation function in sqldf(), need to calculate manually to avoid base R

# calculate mean
sqlStmt <- paste0(
"SELECT AVG(strength) AS 'mean_str' ",
"FROM dbConcrete")
mean_strength <- sqldf::sqldf(sqlStmt)$mean_str

# calculate sample size
sqlStmt <- paste0(
  "SELECT COUNT(strength) as 'num_sample' ",
  "FROM dbConcrete ")
sample <- sqldf::sqldf(sqlStmt)$num_sample

# calculate standard deviation
sqlStmt <- paste0(
  "SELECT SQRT(SUM((strength - ",mean_strength,") * (strength - ",mean_strength, ")) / ",sample," - 1) AS 'stdev_str' ",
  "FROM dbConcrete")
sd_strength <- sqldf::sqldf(sqlStmt)$stdev_str


#cutoff values
outlier_upper <- round(mean_strength + (3 * sd_strength), 2)
outlier_lower <- round(mean_strength - (3 * sd_strength), 2)

#query values of interest
#there are none but this would display all columns for them, test with different numbers
sqlStmt <- paste0(
  "SELECT * ",
  "FROM dbConcrete ",
  "WHERE strength > ", outlier_upper,
  " OR strength < ", outlier_lower, 
  " LIMIT 10; "
)
outliers <- sqldf::sqldf(sqlStmt)
outliers
```

```{r double_check}
# DOUBLE CHECK 
meanR <- mean(dbConcrete$strength)
stdevR <- sd(dbConcrete$strength)
outlier_Rupper <- round((meanR + (3 * stdevR)),2)
outlier_Rlower <- round((meanR - (3 * stdevR)),2)

```
The total sample size = **`r sample`**

Mean Strength was calculated with SQL to be = **`r mean_strength`**

One standard deviation was calculated with SQL to be = **`r sd_strength`**

With outliers defined as 3 standard deviations from the mean, that makes our range **`r outlier_lower`** to **`r outlier_upper`**. There are no outliers in the given dataset. The *sample maximum* value is **`r max(dbConcrete$strength)`** and the *sample minimum* value is **`r min(dbConcrete$strength)`** which is within the range and therefore we don't have outliers.

I also double checked the values with base R calculations for mean = **`r meanR`** and sd = **`r stdevR`**.The numbers are minorly different resulting in slightly different cutoff values but that still does not impact the outliers in the table.
