---
title: "DA5020 Assignment 07.2"
author: "Larkin, Yvonne"
date: "Sp 2025, February 25"
output:
  html_document:
    df_print: paged
---
Instructions said to display results but did not mention about echoing code. I decided to echoed the code, had the object of interest printed as last line of a code chunk, and answered in a sentence with R markdown after each code chunk.

```{r loadXML, warning=FALSE, message=FALSE}
library(XML)
library(RCurl)

#url
xml.url <- getURL("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/Books7.xml")
#load as DOM
book.dom <- xmlParse(xml.url, validate = FALSE, asText = TRUE)
```

### Question 4
Who is the author of the book with the title "Paradox Lost and Found"?
```{r Q4}
# 40min-45min of intro video (video helped a lot for entire assignment)
#APA citation for canvas page: Lecture Video embedded at the Bottom. 07.1 / Overview & Lecture. (2024, February). https://northeastern.instructure.com/courses/206741/pages/07-dot-1-slash-overview-and-lecture?module_item_id=11745756 

# XPath expression by title
# // is relative to look anywhere in file
#want child node title = criteria and retrieve sibling node author
xpathEx <- "//books/book[title='Paradox Lost and Found']/author"

#apply search criteria to extract value of author
author_name <- xpathSApply(book.dom, xpathEx, xmlValue)
author_name
```
The author is `r author_name`.

### Question 5
What are the title of the books written by "Corben, Ellen"?
```{r Q5}
#XPath expression by author

xpathEx <- "//books/book[author='Corben, Ellen']/title"

#apply search criteria to extract value of book title 
titles_byCorben <- xpathSApply(book.dom, xpathEx, xmlValue)
titles_byCorben
```
The title of the books written by "Corben, Ellen" are:
`r titles_byCorben`

### Question 6
How many books are in their second or higher edition?
```{r Q6}
#around 48min of intro video
#APA citation for canvas page: Lecture Video embedded at the Bottom. 07.1 / Overview & Lecture. (2024, February). https://northeastern.instructure.com/courses/206741/pages/07-dot-1-slash-overview-and-lecture?module_item_id=11745756 
#by the way this was mind blowing as an explanation but made sense after
xpathEx <- "count(//books/book[edition>=2]/title)"

#apply search criteria to extract number of books on second or greater edition
num_books <- xpathSApply(book.dom, xpathEx, xmlValue)
num_books
```
There are `r num_books` books in their second or higher edition.

### Question 7
What are the titles of the books that sell for the lowest price?
```{r}
#use min aggregation function
# retrieve prices
xpathEx <- "//books/book[price=min(//books/book/price)]/title"

#apply search criteria to price values
book_lowestprice <- xpathSApply(book.dom, xpathEx, xmlValue)
book_lowestprice
```
The title of the book that sells for the lowest price is `r book_lowestprice`.

### Question8
List the title and prices of all books in the "Romance" or "Science Fiction" genres.
```{r}
#APA citation for help combining expressions: OpenAi. (2025b, February 25). XPath error fix. ChatGPT. https://chatgpt.com/share/67be54e4-e0bc-8006-b930-7632bd4c8e1a
# retrieve book title
xpathEx <- "//books/book[genre='Science Fiction' or genre='Romance']/title"
# retrieve book price
xpathEx2 <- "//books/book[genre='Science Fiction' or genre='Romance']/price"

#apply search criteria to book title with price
books_title <- xpathSApply(book.dom, xpathEx, xmlValue)
books_price <- xpathSApply(book.dom, xpathEx2, xmlValue)
df.display <- data.frame(Title = books_title,
                         Price = books_price,
                         stringsAsFactors = FALSE)
```
The Title and Prices of all the books in either "Romance" or "Science Fiction" are displayed in data frame:
`r df.display`

### Question 9
What is the number of all books that cost $80.00 or more?
```{r}
#around 48min of intro video
#APA citation for canvas page: Lecture Video embedded at the Bottom. 07.1 / Overview & Lecture. (2024, February). https://northeastern.instructure.com/courses/206741/pages/07-dot-1-slash-overview-and-lecture?module_item_id=11745756 
#by the way this was mind blowing as an explanation but made sense after
xpathEx <- "count(//books/book[price>=80.00]/title)"

#apply search criteria to extract number of books on second or greater edition
num_books <- xpathSApply(book.dom, xpathEx, xmlValue)
num_books
```
The number of books that cost more than $80.00 = `r num_books`.

### Question 10
Use a Combination of XPath and R to find the titles and authors of all books or magazines that have "Garden" in their description or genre if there is no description.
```{r Q11}
#'book.dom' (the XML file as DOM) established in first code chunk

#root node
book <- xmlRoot(book.dom)
#root node has 2 child nodes: 1 is books and 2 is magazines
n <- xmlSize(book)

#all books in root node
book_list <- book[[1]]
# 16 books (i.e child nodes)
num_books <- xmlSize(book_list)

#initialize empty df
garden_df <- data.frame(Title = character(),
                          Author = character(),
                          stringsAsFactors = FALSE)
#iterate through books and search their index
for(i in 1:num_books){
  #index genre and description
  genre <- xmlValue(book_list[[i]][[3]])
  description <- xmlValue(book_list[[i]][[7]])
  #if either contain 'Garden append both to garden_df
  if(grepl("Garden", genre) | grepl("Garden", description)){
    #add c(author, title) by index
    author <- xmlValue(book_list[[i]][[1]])
    title <- xmlValue(book_list[[i]][[2]])
    garden_df <- rbind(garden_df, data.frame(Title = title, Author = author, stringsAsFactors = FALSE))
  } 
}

#repeated same process here because it was the first way I thought to solve this, I would need more time to refine further down to one loop
# all magazines in root node
mag_list <- book[[2]]
# 4 magazines (i.e. child nodes)
num_mags <- xmlSize(mag_list)

#iterate through magazines and search their index, but append to same garden_df
for(i in 1:num_mags){
  #index genre only since magazine has no description
  genre <- xmlValue(mag_list[[i]][[3]])
  #if it contains 'Garden append author and title to garden_df
  if(grepl("Garden", genre)){
    #add author and title by index
    author <- xmlValue(mag_list[[i]][[1]])
    title <- xmlValue(mag_list[[i]][[2]])
    garden_df <- rbind(garden_df, data.frame(Title = title, Author = author, stringsAsFactors = FALSE))
  } 
}
```
All Books or Magazines pertaining to "Garden" are displayed with the Title of Books and their Author in data frame:
`r garden_df`


### Question 11
Use a combination of XPath and R to find the median price of all magazines published in 2023
```{r}
xpathEx <- "//magazines/magazine[year=2023]/price"

#all magazine prices for 2023
mag_price <- xpathSApply(book.dom, xpathEx, xmlValue)
#place prices in order and make numeric
sorted_price <- sort(as.numeric(mag_price))
#find median
median_price <- median(sorted_price)
median_price
```
The median price of all magazines published in 2023 is: `r median_price` (the units are not defined).
